---
title: 'EDA, Sample Selection Criteria, Descriptive Statistics'
author: "Alan, Free, and Mai"
date: "02/04/2020"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include = FALSE}
## Course: ECON 5300
## Purpose: Assignment 3 - Preliminary Economic Estimates
## Date: January 28, 2020
## Authors: Alan, Free, and Mai

```

```{r clear env., message = FALSE, echo = FALSE}
# Clear environment of variables and functions
rm(list = ls(all = TRUE))

# Clear environment of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

# Assignment Description

+ LMA Assignment 3 - Preliminary Economic Estimates

+ Research Questions:

    - How do earnings vary by marital status?
    - Does the marriage premium or penalty vary by gender?


# Load Libraries

```{r libraries, message = FALSE}
# Load libraries
library(tidyverse)
library(stargazer)
library(psych)
library(corrplot)
library(lmtest) #BGP test
library(robustbase)
library(sandwich) # robust SE

```


#Load and transform data

```{r load data}
# Load data
df <- read.csv("pop_wa.csv", header = TRUE) 

# create a numeric column for married status
df <- df %>%
  mutate(Married = ifelse(MAR == 1, 1, 0))

# create a new column --unearned income
df <- df %>%
  mutate(UI = PINCP - PERNP)
         
```

```{r transform dummy variables}
# Educational Attainment
df$NoDegree <- ifelse(df$SCHL < 15, 1, 0) 
df$HighSchool <- ifelse(df$SCHL > 14 & df$SCHL < 20 , 1, 0) # contains GED, some college
df$CollegeDegree <-ifelse(df$SCHL == 20 | df$SCHL == 21, 1, 0) # contains Associates and Bachelors
df$HigherEducation <- ifelse(df$SCHL > 21, 1, 0) # contains masters and beyond

# Race/Ethnicity
df$White <- ifelse(df$RAC1P == 1, 1, 0)
df$Black <- ifelse(df$RAC1P == 2, 1, 0)
df$Asian <- ifelse(df$RAC1P == 6, 1, 0) # only 7.9% asian- I think that's enough?
df$OtherRace <- ifelse(df$RAC1P == 3 | df$RAC1P == 4 | df$RAC1P == 7 | df$RAC1P == 8 | df$RAC1P == 9 | df$RAC1P == 5, 1, 0) 
# other contains AmericanIndian, Alaska, Hawaii, OtherRace, MultipleRace, and AI.AN

# recode sex to make female = 1
df$Female <- ifelse(df$SEX == 2, 1, 0)

# create age-squared
df$AgeSQ <- df$AGEP * df$AGEP
df$Age <- df$AGEP

# interaction term with female and marriage
df$Female.Married = df$Female * df$Married


```

## implementing selection criteria
```{r}
# we only want incomes, earnings, and wages to be positive so we will exclude anything less than or equal to 0 
df_ear <- df[which(df$PERNP >= 20020 & (df$WKW == 1 | df$WKW == 2 | df$WKW == 3) & df$WKHP >= 35), ]

```

## Descriptive Statistics
```{r}
# change earnings to a better name for paper
df_ear$Earnings <- df_ear$PERNP

df <- df_ear %>%
  select(Earnings, NoDegree, HighSchool, CollegeDegree, HigherEducation, White, Black, Asian, OtherRace, Female, Married, Female.Married, Age, AgeSQ)

# uni-variate non-graphical EDA
stargazer(df, type = "text", header = FALSE, median = TRUE)

```

+ creating a new column named "MARRIED" 
 
    - to separate a person is now "married" from others such as separated, divorced, under 15 years old, etc. 

+ SEX  
 
    - male and female (1 and 2)
    - the sample has almost equal proportions of males and females. Females being 50.6% of the sample 
    
+ MARRIED  
 
    - 44.2% of individuals are married, or "now married" 
    - Others include divorced, separated, person under 15 years old, etc..
    
+ WAGP 
 
    - wages or salary income in the past 12 months
    - average wages in the past 12 months was $32,951.53
    - the median wages/salary income past 12 months was $10,000
    - 25% individuals have had 0 wages/salary income
    - individuals in this group could be in the under 15 years age group. Or they didn't earn any money but someone else in their household did
    - There are large number of missing values (74695-61871) 12,824 which may explains individuals with zero wages
    
+ PERNP 
 
    - total person's earnings
    - average total person's earnings is $35,570.04
    - the median total person's earnings is $15,000
    - as same as wages in the past 12 months, PERNP has a big gap between Median and Maximum, $821,000. These are likely outliers. 
    - there are a big number of missing values, might explain the 25% individuals with 0 earnings
    
+ PINCP
 
    - total person's income
    - average total person's income is $45,619.32
    - the median total person's income is $28,600
    - potentially have outliers with a maximum value of $1,038,000
    - the minimum total person's income is -$5,100 (loss)
    - 25% individuals have had a total income of $9,300, this could come from the unearned income 
    
+ UI
 
    - a new column created "Unearned Income", calculated by substrating total person's income by total person's earnings
    - unearned income: interest, dividends and transfer payments
    - there are obviously missing values due to missing values of PERNP and PINCP
    - 25% - 50$ individuals do not obtain unearned income
    - as a consequence of a large maximum value of total person's earnings and total person's income, the maximum of unearned income is $401,000
    
    
# Base EDA

```{r view data}
# Generate histogram for select variable(s)
# marital status
par(mfrow = c(2,2))
  hist(df$MAR,col = "orange")
  hist(df$MARRIED, col = "orange")
  hist(df$PERNP, col = "orange")
  hist(df$SEX, col = "orange")
```

+ Observations verified and confirmed from the above analysis

    - even distribution between males and females
    - near even distribution of married couples vs non-married
    - earnings is skewed with a long right tail, therefore median is the best measure
    

```{r boxplot}
# Generate boxplot for select variable(s)
# marital status
par(mfrow = c(2,2))
  boxplot(df$MAR,main = "MARITAL STATUS",col = "orange")
  boxplot(df$MARRIED,main = "MARRIED", col = "orange")
  boxplot(df$PERNP, main = "Earnings", col = "orange")
  boxplot(df$SEX,main = "GENDER", col = "orange")
```

+ different view of the same picture we got from the histograms


```{r barplox}
# Generate barplot for select variable(s)
# marital status
par(mfrow = c(2,2))
  barplot(table(df$MAR),main = "MARITAL STATUS",col = "orange")
  barplot(table(df$MARRIED),main = "MARRIED", col = "orange")
  barplot(table(df$SEX), main = "GENDER", col = "orange")
  
```

+ MARRIED
 
    - slighly more non-married than married
    
+ GENDER
 
    - As seen before, almost equal distribution between male and female
    
+ MARITAL STATUS
 
    - "married" seems to have the highest proportion
    - though, its total is lower than other categories combined
    

```{r correlation}
df %>%
  select(MARRIED,PERNP,PINCP) %>%
  cor(use = "pairwise.complete.obs", method = "pearson") %>%
  corrplot(type = "full", method = "number", diag = FALSE)

```

+ CORRELATION
 
    - Strong correlation between total person's income and total person's earnings as expected
    - Low correlation between being married or not and total person's income/total person's earnings
    - should not be a big concern in this case
    

```{r geom_point}
# how does marital status look with earnings)
df %>%
  select(MAR,PERNP) %>%
  ggplot(aes(x = MAR, y = PERNP)) + 
  geom_point(color = "orange") + 
  theme_minimal()

```

+ Married person seems to earn the most as expected


```{r geom}
# how do marital status and earnings look with gender)
df %>%
  select(MAR,PERNP,SEX) %>%
  ggplot(aes(x = MAR, y = PERNP, color = SEX)) + 
  geom_point() + 
  theme_minimal()

```

+ Across all type of marital status, males earn more than females


# Economic Estimates

## Build models based on theoretical grounds
```{r}
# using PERNP as the dependent variable (Left out noDegree, Other (race) to avoid perfect multicollinearity)
estimate_earnings <- lm(Earnings ~ HighSchool + CollegeDegree + HigherEducation + White + Black + Asian + Married + Female + Female.Married + Age + AgeSQ , data = df)

summary(estimate_earnings)

```

+ Age +, Age^2 - 

    - this means that as people get older, the effect of age on earnings gets weaker
  
```{r}
# graphing the regression
par(mfrow = c(2, 2)) # init 4 charts in 1 panel
plot(estimate_earnings)

```

+ as seen in the upper left, residuals seem to increase as the fitted y value increases

    - the inference here is that heteroskedasticity exists
    

```{r}
# Breusch-Pagan test
bptest(estimate_earnings)

```

+ our p-value is statistically significant at the 0.05 level

    - this means we can reject the null hypothesis that the variance of the residuals is constant and heteroskedasticity 
     is indeed present, confirming the results in the graphical representation

+ we need to rectify this, so we will make an attempt by using log earnings


```{r}
# log function and bp test for log earnings equation
estimate_logearnings = lm(log(Earnings) ~ HighSchool + CollegeDegree + HigherEducation + White + Black + Asian + Married + Female + Female.Married + Age + AgeSQ, data = df)

summary(estimate_logearnings)

bptest(estimate_logearnings)

```

+ according to the Breusch-Pagan test on the log earnings model, we still have heteroskedasticity

    - need to calculate robust standard errors

## Subsetting Men and Women - Married and not Married
```{r}
# Men - Married
df_MM <- df[which(df$Female == 0 & df$Married == 1), ]
model_MM <- lm(log(Earnings) ~ HighSchool + CollegeDegree + HigherEducation + White + Black + Asian + Age + AgeSQ, data = df_MM)
summary(model_MM)

vcv <- vcovHC(model_MM, type = "HC3")
coeftest(model_MM, vcv)

stargazer(df_MM, type = "text", header = FALSE, median = TRUE)
```

```{r}
# Men - Unmarried
df_MU <- df[which(df$Female == 0 & df$Married == 0), ]
model_MU <- lm(log(Earnings) ~ HighSchool + CollegeDegree + HigherEducation + White + Black + Asian + Age + AgeSQ, data = df_MU)
summary(model_MU)

vcv <- vcovHC(model_MU, type = "HC3")
coeftest(model_MU, vcv)

stargazer(df_MU, type = "text", header = FALSE, median = TRUE)
```

```{r}
# Women - Married
df_WM <- df[which(df$Female == 1 & df$Married == 1), ]
model_WM <- lm(log(Earnings) ~ HighSchool + CollegeDegree + HigherEducation + White + Black + Asian + Age + AgeSQ, data = df_WM)
summary(model_WM)

vcv <- vcovHC(model_WM, type = "HC3")
coeftest(model_WM, vcv)

stargazer(df_WM, type = "text", header = FALSE, median = TRUE)
```

```{r}
# Women - Unmarried
df_WU <- df[which(df$Female == 1 & df$Married == 0), ]
model_WU <- lm(log(Earnings) ~ HighSchool + CollegeDegree + HigherEducation + White + Black + Asian + Age + AgeSQ, data = df_WU)
summary(model_WU)

vcv <- vcovHC(model_WU, type = "HC3")
coeftest(model_WU, vcv)

stargazer(df_WU, type = "text", header = FALSE, median = TRUE)
```


## Robust SEs
```{r}
# using sandwich library (HC3 is the most conservative correction)
vcv <- vcovHC(estimate_logearnings, type = "HC3")

# robust t-tests
coeftest(estimate_logearnings, vcv)

```

+ I use HC3 because I am unsure which version (0, 1, 2, or 3) is the best

    - We get our adjusted standard errors and output




## -- confused from here on lol ~


```{r}
#submodel by race
White <- df[which(df$White ==1), ]
Black <- df[which(df$Black ==1), ]
Asian <- df[which(df$Asian ==1), ]
Other <-df[which(df$Other ==1), ]

LogEarnings.white <- lm(log(PERNP) ~ HighSchool + CollegeDegree + HigherEducation + MARRIED + SEX + PINCP + WAGP + AGESQ, data = White)
LogEarnings.asian <- lm(log(PERNP) ~ HighSchool + CollegeDegree + HigherEducation + MARRIED + SEX + PINCP + WAGP + AGESQ, data = Asian)
LogEarnings.black <- lm(log(PERNP) ~ HighSchool + CollegeDegree + HigherEducation + MARRIED + SEX + PINCP + WAGP + AGESQ, data = Black)
LogEarnings.Other <-lm(log(PERNP) ~ HighSchool + CollegeDegree + HigherEducation + MARRIED + SEX + PINCP + WAGP + AGESQ, data = Other)
```


```{r}
#BP Testing
bptest(LogEarnings.white)
bptest(LogEarnings.black)
bptest(LogEarnings.asian)
bptest(LogEarnings.Other)
```

```{r}
#Calculate robus SE (White)
white.robus.se<sqrt(diag(vcovHC(LogEarnings.white, type = "HC")))
coeftest(LogEarnings.white, vcov = vcovHC(LogEarnings.white, type = "HC"))
as.data.frame(white.robus.se)
```

```{r}
#Calculate robus SE (Black)
black.robus.se<sqrt(diag(vcovHC(LogEarnings.black, type = "HC")))
coeftest(LogEarnings.black, vcov = vcovHC(LogEarnings.black, type = "HC"))
as.data.frame(black.robus.se)
```

```{r}
#Calculate robus SE (asian)
asian.robus.se<sqrt(diag(vcovHC(LogEarnings.asian, type = "HC")))
coeftest(LogEarnings.asian, vcov = vcovHC(LogEarnings.asian, type = "HC"))
as.data.frame(asian.robus.se)
```

```{r}
#Calculate robus SE (other)
other.robus.se<sqrt(diag(vcovHC(LogEarnings.Other, type = "HC")))
coeftest(LogEarnings.Other, vcov = vcovHC(LogEarnings.Other, type = "HC"))
as.data.frame(other.robus.se)
```






























